---
title: "Differential Gene Expression Analysis"
author: "RNAseq Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# snakemake <- readRDS(".deg.Rmd.RDS")
saveRDS(snakemake, ".deg.Rmd.RDS")
library(edgeR)
library(data.table)
library(DESeq2)
library(ggplot2)
library(reshape2)
library(pheatmap)
library(RColorBrewer)
library(dplyr)
library(tibble)
```

## Load counts and samplesheet
```{r load-counts, include=TRUE}
# Load teacher-provided count matrix and subset to our 6 samples
count_matrix <- readRDS(snakemake@input$count_matrix)
our_samples  <- snakemake@params$samples
count_matrix <- count_matrix[, colnames(count_matrix) %in% our_samples]
cat("Dimensions after subsetting:", dim(count_matrix), "\n")
head(count_matrix)
```
```{r load-samplesheet, include=TRUE}
samplesheet <- fread(snakemake@input$samplesheet)
samplesheet$phenotype <- relevel(factor(samplesheet$phenotype), ref = "WT")
# DESeq2 requires rownames of colData to match colnames of count matrix
samplesheet <- as.data.frame(samplesheet)
rownames(samplesheet) <- samplesheet$sample
head(samplesheet)
```

## Build DESeq2 object
```{r deseq-object}
dds <- DESeqDataSetFromMatrix(count_matrix,
                              colData = samplesheet,
                              design = ~ phenotype)
dds
```

## Filter low-count genes
```{r filter}
keep <- rowSums(counts(dds) >= 10) >= 2
dds_filt <- dds[keep, ]
dds_filt
```

## Exploratory visualisation

### Raw count distribution
```{r boxplot-raw, include=TRUE}
melted_counts <- reshape2::melt(as.matrix(counts(dds_filt)),
                                varnames = c("Gene", "ID"),
                                value.name = "Count")
ggplot(melted_counts, aes(x = ID, y = Count)) +
  geom_boxplot(outlier.size = 0.2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Raw count distribution per sample")
```

### VST-normalised count distribution
```{r vst, include=TRUE}
vsd <- vst(dds_filt, blind = FALSE)
vsd_counts <- assay(vsd)

melted_vsd <- reshape2::melt(as.matrix(vsd_counts),
                             varnames = c("Gene", "ID"),
                             value.name = "Count")
ggplot(melted_vsd, aes(x = ID, y = Count)) +
  geom_boxplot(outlier.size = 0.2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("VST-normalised count distribution per sample")
```

### Dispersion estimates
```{r dispersion, include=TRUE}
dds_filt <- estimateSizeFactors(dds_filt)
dds_filt <- estimateDispersions(dds_filt)
plotDispEsts(dds_filt)
```

### PCA
```{r pca, include=TRUE}
plotPCA(vsd, intgroup = "phenotype") +
  theme_minimal() +
  ggtitle("PCA of VST counts")
```

### Heatmap – top 500 most variable genes
```{r heatmap-variable, include=TRUE}
vars <- apply(assay(vsd), 1, var)
top <- names(sort(vars, decreasing = TRUE))[1:500]
mat <- assay(vsd)[top, ]
ann_col <- as.data.frame(colData(vsd)[, "phenotype", drop = FALSE])
pheatmap::pheatmap(
  mat,
  scale = "row",
  show_rownames = FALSE,
  annotation_col = ann_col,
  main = "Top 500 most variable genes",
  cluster_cols = FALSE
)
```

### Sample distance heatmap
```{r heatmap-dist, include=TRUE}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$sample, vsd$phenotype, sep = "-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```

## Differential gene expression
```{r deseq-run, include=TRUE}
dds_filt <- DESeq(dds_filt)
res <- results(dds_filt, contrast = c("phenotype", "WT", "CHD8_KO"))
summary(res)
```
```{r lfc-shrink}
resLFC <- lfcShrink(dds_filt,
                    coef = "phenotype_CHD8_KO_vs_WT",
                    type = "apeglm")
resLFC
```

### Top 10 DEGs
```{r top10, include=TRUE}
top_genes <- head(resLFC[order(resLFC$padj), ], n = 10)
top_genes
```

### Top DEG counts per sample
```{r top-counts, include=TRUE}
top_genes_names <- rownames(top_genes)
par(mfrow = c(3, 3))
for (g in top_genes_names[1:9]) {
  plotCounts(dds_filt, gene = g, intgroup = "phenotype")
}
par(mfrow = c(1, 1))
```

### Volcano plot
```{r volcano, include=TRUE}
with(res, plot(log2FoldChange, -log10(pvalue),
               pch = 20, main = "Volcano plot", xlim = c(-3, 3)))
with(subset(res, padj < .05),
     points(log2FoldChange, -log10(pvalue), pch = 20, col = "blue"))
with(subset(res, padj < .05 & abs(log2FoldChange) > 2),
     points(log2FoldChange, -log10(pvalue), pch = 20, col = "red"))
```

## Gene Set Enrichment Analysis (GSEA)
```{r gsea, include=TRUE}
library(fgsea)
library(msigdbr)

msig_mouse <- msigdbr::msigdbr(species = "Mus musculus", category = "C5") %>%
  dplyr::select(gs_name, ensembl_gene) %>%
  dplyr::filter(!is.na(ensembl_gene), ensembl_gene != "") %>%
  dplyr::distinct()

pathways <- split(msig_mouse$ensembl_gene, msig_mouse$gs_name)

rank_tbl <- cbind(ENSEMBL = rownames(as.data.frame(resLFC)),
                  as.data.frame(resLFC))
ranks <- rank_tbl$log2FoldChange
names(ranks) <- rank_tbl$ENSEMBL
ranks <- sort(ranks, decreasing = TRUE)

set.seed(1)
fg <- fgsea(
  pathways = pathways,
  stats    = ranks,
  minSize  = 15,
  maxSize  = 500
) %>% as_tibble() %>% arrange(padj)

head(fg[, c("pathway", "NES", "pval", "padj")], 20)
```

### Enrichment plots – highest and lowest NES
```{r gsea-plots, include=TRUE}
high_nes <- fg[order(-fg$NES), ]$pathway[1]
plotEnrichment(pathways[[high_nes]], ranks) + labs(title = high_nes)

low_nes <- fg[order(fg$NES), ]$pathway[1]
plotEnrichment(pathways[[low_nes]], ranks) + labs(title = low_nes)
```

### Top 10 up- and down-regulated pathways
```{r gsea-dotplot, include=TRUE, fig.width=12, fig.height=6}
top_pos <- fg %>% filter(NES > 0) %>% slice_max(NES, n = 10)
top_neg <- fg %>% filter(NES < 0) %>% slice_min(NES, n = 10)

top_pathways <- bind_rows(top_pos, top_neg) %>% arrange(NES)

ggplot(top_pathways, aes(x = NES,
                          y = reorder(pathway, NES),
                          size = -log10(padj),
                          color = NES > 0)) +
  geom_point() +
  scale_color_manual(values = c("red", "blue"),
                     labels = c("Downregulated", "Upregulated")) +
  labs(x = "Normalized Enrichment Score (NES)",
       y = "Pathway",
       size = "-log10(padj)") +
  theme_minimal() +
  theme(legend.title = element_blank(),
        axis.text.y  = element_text(size = 7))
```
